{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}

### 2016-08-24 ident_manu.Rnw
table 1 do sens all cats added

\noindent where the estimate depended on the change in the predicted value for response variable $y$ divided by the change in the parameter $\Theta_j$ multiplied by the quotient of scaling factors $w$ for each.  The scaling factors, $w_{\Theta_j}$ for the parameter $\Theta_j$ and $w_{y_i}$ for response variable $y_i$, were set as the default value of the unperturbed parameter and the predicted value of $y_i$ after perturbation \citep{Soetaert10}.  The scaling ensures the estimates are unitless such that the relative magnitudes allows comparisons of model sensitivity to parameter changes and state variables that differ in scale.  Estimates for $S_{ij}$ were summarized as $L1$ and $L2$ across the time series to obtain individual sensitivity values of \ac{do} in response to a change in parame ter $j$:

\begin{equation} \label{l1}
L1 = \sum|S_{ij}|/n
\end{equation}
\begin{equation} \label{l2}
L2 = \sqrt{\sum\left(S_{ij}^2\right)|/n}
\end{equation}

In general, positive sensitivity estimates suggested a parameter had a positive effect on \ac{do} for a given increase in the parameter, whereas the converse was true for negative sensitivity estimates.  However, the effect of a parameter change may not be uniform over time such that $S_{ij}$ can change in magnitude and sign depending on the temporal location.  Time series of \ac{do} estimates before and after perturbation were also evaluated to identify patterns not captured by the summary statistics. All parameters for each of the six equation categories (optics, organic matter, phytoplankton, temperature, and zooplankton) that had non-zero $L1$ or $L2$ were retained for identifiability analysis.

@@

As a final analysis, the effects of observational and structural uncertainty on the sensitivity analyses were evaluated for \ac{do}.  These analyses used changes in parameter uncertainty and identifiability to infer the effects of observational and structural uncertainty on model output to create a more comprehensive evaluation.  First, observational uncertainty was evaluated by varying the initial conditions supplied to the model.  The initial conditions were based on observational data from research cruises in the northern \ac{gom} \citep{Murrell14} and uncertainty in these data translates directly to uncertainty that can influence results of the sensitivity analysis.  For example, the sensitivity of \ac{do} to variation in the half-saturation constants (the concentration supporting half the maximum uptake rate of nutrients) will vary given the initial nutrient concentrations \citep{Eppley69}.  Further, changes in the ratio between nitrogen and phosphorus could affect sensitivity depending on which nutrient is limiting .  Parameter sensitivity and identifiability was re-evaluated for \ac{do} by varying all initial conditions that were non-zero by 50\% and 200\% of the original values that were estimated from the observed data.

@@

% optics sensitivity
<<optsens, eval = F, cache = F>>=
senstab1('Optics', 'optsens')
@

% organic matter sensitivity
<<omsens, eval = F, cache = F>>=
senstab1('Organic Matter', 'omsens')
@

% phytoplankton sensitivity
<<phytosens, eval = F, cache = F>>=
# data(sens_ests_cat)
# tmp <- filter(sens_ests_cat, Category == 'Phytoplankton')
# quantile(tmp$error)
senstab2('Phytoplankton', 'phytosens', sub = 0.59, sub.txt = 'Parameters less than the 75\\textsuperscript{th} percentile (0.59) for L1 were removed for brevity.')
@

% temperature sensitivity
<<tempsens, eval = F, cache = F >>=
senstab2('Temperature', 'tempsens')
@

% zooplankton sensitivity
<<zoopsens, eval = F, cache = F>>=
senstab2('Zooplankton', 'zoopsens')
@

@@

<<sensalltile, eval = T, echo = F, fig.height = 7, fig.width = 6, out.width = '\\textwidth', fig.cap = 'Sensitivity values (L1, \\cref{l1}) for local analyses of all state variables. Parameters are grouped by category: optics, organic matter, phytoplankton, zooplankton, temperature, and zoplankton.'>>=
data(sens_ests_cat_all)

>>>>

\noindent where the estimate depended on the change in the predicted value for response variable $y$ divided by the change in the parameter $\Theta_j$ multiplied by the quotient of scaling factors $w$ for each.  The scaling factors, $w_{\Theta_j}$ for the parameter $\Theta_j$ and $w_{y_i}$ for response variable $y_i$, were set as the default value of the unperturbed parameter and the predicted value of $y_i$ after perturbation \citep{Soetaert10}.  The scaling ensures the estimates are unitless such that the relative magnitudes allows comparisons of model sensitivity to parameter changes and state variables that differ in scale.  Estimates for $S_{ij}$ were summarized as $L1$ to obtain individual sensitivity values of \ac{do} in response to a change in parame ter $j$:

\begin{equation} \label{l1}
L1 = \sum|S_{ij}|/n
\end{equation}

In general, positive sensitivity estimates suggested a parameter had a positive effect on \ac{do} for a given increase in the parameter, whereas the converse was true for negative sensitivity estimates.  However, the effect of a parameter change may not be uniform over time such that $S_{ij}$ can change in magnitude and sign depending on the temporal location.  Time series of \ac{do} estimates before and after perturbation were also evaluated to identify patterns not captured by the summary statistics. All parameters for each of the six equation categories (optics, organic matter, phytoplankton, temperature, and zooplankton) that had non-zero $L1$ were retained for identifiability analysis.

@@

As a final analysis, the effects of observational and structural uncertainty on the sensitivity analyses were evaluated for \ac{do}.  These analyses used changes in parameter uncertainty and identifiability to infer the effects of observational and structural uncertainty on model output to create a more comprehensive evaluation.  First, observational uncertainty was evaluated by varying the initial conditions supplied to the model.  The initial conditions were based on observational data from research cruises in the northern \ac{gom} \citep{Murrell14} and uncertainty in these data translates directly to uncertainty that can influence results of the sensitivity analysis.  For example, the sensitivity of \ac{do} to variation in the half-saturation constants (the concentration supporting half the maximum uptake rate of nutrients) will vary given the initial nutrient concentrations \citep{Eppley69}.  Further, changes in the ratio between nitrogen and phosphorus could affect sensitivity depending on which nutrient is limiting.  Parameter sensitivity and identifiability was re-evaluated for \ac{do} by varying all initial conditions that were non-zero by 50\% and 200\% of the original values that were estimated from the observed data.

@@

<<dosens, eval = T, cache = F>>=
load(file = 'data/sens_ests_cat_all.RData')
load(file = 'data/sens_ests_all.RData')

# subset the sensitivy and sensitivity catgory lists
tmpest <- sens_ests_all[['O2']]
tmpest_cat <- sens_ests_cat_all[['O2']]

labs <- parcats2(as_df = TRUE) %>% 
  rename(Parameter = shrt)

totab <- tmpest_cat %>% 
  select(Parameter) %>% 
  left_join(., tmpest$sens, by = 'Parameter') %>% 
  left_join(., labs, by = 'Parameter') %>% 
  select(cats, lngs, Parameter, vals, L1) %>% 
  arrange(cats, -L1) %>% 
  mutate(
    Parameter = par_txt(Parameter),
    lngs = gsub('^.*:\\s', '', lngs), 
    lngs = gsub('_', '', lngs), 
    vals = sapply(vals, scinot),
    L1 = sapply(L1, scinot, digits = 2, pow = 2)
  ) %>% 
  rename(
    Description = lngs, 
    Value = vals
  )

# final table formatting
Description <- totab$Description
cats <- totab$cats
totab <- totab[,-c(1:2)]

cap.val<- paste('Sensitivities of \\ac{do} to perturbation of indivividual parameters.  Sensitivities are based on a 50\\% increase from the initial parameter value, where $L1$ summarizes differences in model output from the default (see \\cref{l1}).  Parameters that did not affect \\ac{do} are not shown.  Parameters are grouped by categories as optics, temperature, phytoplankton, zooplankton, and organic matter.')

latex( 
  totab,
  file = '',
  rowlabel = 'Description',
  caption = cap.val,
  caption.loc = 'top',
  rowname = Description,
  rgroup = unique(cats),
  n.rgroup = as.numeric(table(cats)),
  size = 'footnotesize',
  label = 'tab:dosens'
  )
@

@@

<<sensalltile, eval = T, echo = F, fig.height = 7, fig.width = 6, out.width = '\\textwidth', fig.cap = 'Sensitivity values (L1, \\cref{l1}) for local analyses of all state variables. Parameters are grouped by category: optics, organic matter, phytoplankton, zooplankton, temperature, and zoplankton.  See \\cref{tab:dosens} for L1 values for \\ac{do}.'>>=
data(sens_ests_cat_all)

### 2016-08-24 ident_manu.tex
table 1 do sens all cats added

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$, Lisa L. Lowe$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
	\\\\{\textit {\normalsize $^2$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Tue Aug 23 17:18:33 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

As a final analysis, the effects of observational and structural uncertainty on the sensitivity analyses were evaluated for \ac{do}.  These analyses used changes in parameter uncertainty and identifiability to infer the effects of observational and structural uncertainty on model output to create a more comprehensive evaluation.  First, observational uncertainty was evaluated by varying the initial conditions supplied to the model.  The initial conditions were based on observational data from research cruises in the northern \ac{gom} \citep{Murrell14} and uncertainty in these data translates directly to uncertainty that can influence results of the sensitivity analysis.  For example, the sensitivity of \ac{do} to variation in the half-saturation constants (the concentration supporting half the maximum uptake rate of nutrients) will vary given the initial nutrient concentrations \citep{Eppley69}.  Further, changes in the ratio between nitrogen and phosphorus could affect sensitivity depending on the limiting nutrient .  Parameter sensitivity and identifiability was re-evaluated for \ac{do} by varying all initial conditions that were non-zero by 50\% and 200\% of the original values estimated from the observed data.

@@

% optics sensitivity

% organic matter sensitivity

% phytoplankton sensitivity

% temperature sensitivity

% zooplankton sensitivity

@@

\caption[Sensitivity values (L1, \cref{l1}) for local analyses of all state variables]{Sensitivity values (L1, \cref{l1}) for local analyses of all state variables. Parameters are grouped by category: optics, organic matter, phytoplankton, zooplankton, temperature, and zoplankton.}\label{fig:sensalltile}
\end{figure}

>>>>

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$, Lisa L. Lowe$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
	\\\\{\textit {\normalsize $^2$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Wed Aug 24 10:41:50 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

As a final analysis, the effects of observational and structural uncertainty on the sensitivity analyses were evaluated for \ac{do}.  These analyses used changes in parameter uncertainty and identifiability to infer the effects of observational and structural uncertainty on model output to create a more comprehensive evaluation.  First, observational uncertainty was evaluated by varying the initial conditions supplied to the model.  The initial conditions were based on observational data from research cruises in the northern \ac{gom} \citep{Murrell14} and uncertainty in these data translates directly to uncertainty that can influence results of the sensitivity analysis.  For example, the sensitivity of \ac{do} to variation in the half-saturation constants (the concentration supporting half the maximum uptake rate of nutrients) will vary given the initial nutrient concentrations \citep{Eppley69}.  Further, changes in the ratio between nitrogen and phosphorus could affect sensitivity depending on which nutrient is limiting.  Parameter sensitivity and identifiability was re-evaluated for \ac{do} by varying all initial conditions that were non-zero by 50\% and 200\% of the original values that were estimated from the observed data.

@@

%latex.default(totab, file = "", rowlabel = "Description", caption = cap.val,     caption.loc = "top", rowname = Description, rgroup = unique(cats),     n.rgroup = as.numeric(table(cats)), size = "footnotesize",     label = "tab:dosens")%
\begin{table}[!tbp]
{\footnotesize
\caption{Sensitivities of \ac{do} to perturbation of indivividual parameters.  Sensitivities are based on a 50\% increase from the initial parameter value, where $L1$ summarizes differences in model output from the default (see \cref{l1}).  Parameters that did not affect \ac{do} are not shown.  Parameters are grouped by categories as optics, temperature, phytoplankton, zooplankton, and organic matter.\label{tab:dosens}} 
\begin{center}
\begin{tabular}{llll}
\hline\hline
\multicolumn{1}{l}{Description}&\multicolumn{1}{c}{Parameter}&\multicolumn{1}{c}{Value}&\multicolumn{1}{c}{L1}\tabularnewline
\hline
{\bfseries Optics}&&&\tabularnewline
~~Chla specific absorption at 490 nm&\textit{astar490}&$0.04$&$6.27\times 10^{-4}$\tabularnewline
~~OMA specific absorption at 490 nm&\textit{astarOMA}&$0.1$&$3.22\times 10^{-4}$\tabularnewline
~~OMZ specific absorption at 490 nm&\textit{astarOMZ}&$0.1$&$5.35\times 10^{-6}$\tabularnewline
\hline
{\bfseries Temperature}&&&\tabularnewline
~~Optimum temperature for growth(C)&\textit{Tref(nospA+nospZ)}&$22$&$1.8\times 10^{-4}$\tabularnewline
\hline
{\bfseries Phytoplankton}&&&\tabularnewline
~~phytoplankton carbon/cell&\textit{Qc}&$1.35\times 10^{-6}$&$0.71$\tabularnewline
~~initial slope of the photosynthesis-irradiance relationship&\textit{alpha}&$8.42\times 10^{-17}$&$0.03$\tabularnewline
~~mortality coefficient&\textit{mA}&$0.1$&$0.03$\tabularnewline
~~N-uptake rate measured at umax&\textit{vmaxN}&$4.1\times 10^{-8}$&$0.03$\tabularnewline
~~coefficient for non-limiting nutrient&\textit{aN}&$1$&$0.01$\tabularnewline
~~phytoplankton basal respiration coefficient&\textit{respb}&$0.02$&$0.01$\tabularnewline
~~half-saturation constant for P&\textit{Kp}&$2.86$&$0.01$\tabularnewline
~~minimum N cell-quota&\textit{QminN}&$6.08\times 10^{-9}$&$7.44\times 10^{-3}$\tabularnewline
~~maximum growth rate&\textit{umax}&$0.41$&$5.34\times 10^{-3}$\tabularnewline
~~half-saturation constant for N&\textit{Kn}&$4.51$&$1.56\times 10^{-3}$\tabularnewline
~~P-uptake rate measured at umax&\textit{vmaxP}&$2.68\times 10^{-8}$&$1.3\times 10^{-3}$\tabularnewline
~~minimum P cell-quota&\textit{QminP}&$6.19\times 10^{-10}$&$1.59\times 10^{-4}$\tabularnewline
~~phytoplankton growth respiration coefficient&\textit{respg}&$0.1$&$7.63\times 10^{-5}$\tabularnewline
\hline
{\bfseries Zooplankton}&&&\tabularnewline
~~zooplankton carbon/individual&\textit{ZQc}&$3.13\times 10^{-4}$&$4.51\times 10^{-3}$\tabularnewline
~~zooplankton nitrogen/individual&\textit{ZQn}&$6.95\times 10^{-5}$&$1.97\times 10^{-3}$\tabularnewline
~~Zooplankton biomass-dependent respiration factor&\textit{Zrespb}&$0.1$&$1.47\times 10^{-4}$\tabularnewline
~~Zooplankton mortality constant for quadratic mortality&\textit{Zm}&$7.2\times 10^{-4}$&$4.61\times 10^{-5}$\tabularnewline
~~zooplankton phosphorus/individual&\textit{ZQp}&$3.77\times 10^{-6}$&$2.89\times 10^{-5}$\tabularnewline
\hline
{\bfseries Organic Matter}&&&\tabularnewline
~~O2 concentration that inhibits denitrification&\textit{KstarO2}&$10$&$0.05$\tabularnewline
~~turnover rate for OM1A and OM1G&\textit{KG1}&$50$&$0.01$\tabularnewline
~~turnover rate for OM2A and OM2G&\textit{KG2}&$50$&$7.53\times 10^{-3}$\tabularnewline
~~half-saturation concentration for O2 utilization&\textit{KO2}&$10$&$6.49\times 10^{-3}$\tabularnewline
~~half-saturation concentration for NO3 used in denitrification&\textit{KNO3}&$10$&$6.36\times 10^{-3}$\tabularnewline
~~decay rate of CDOM, 1/day&\textit{KGcdom}&$0.01$&$8\times 10^{-4}$\tabularnewline
~~NH4 rate constant for nitrification&\textit{KNH4}&$1$&$2.9\times 10^{-4}$\tabularnewline
~~maximum rate of nitrification per day&\textit{nitmax}&$0.52$&$2.44\times 10^{-4}$\tabularnewline
\hline
\end{tabular}\end{center}}

\end{table}

@@

\caption{Sensitivity values (L1, \cref{l1}) for local analyses of all state variables. Parameters are grouped by category: optics, organic matter, phytoplankton, zooplankton, temperature, and zoplankton.  See \cref{tab:dosens} for L1 values for \ac{do}.}\label{fig:sensalltile}
\end{figure}
